Redundant calibration as a complex optimization problem

### 2017-11-03 paper_G_final.tex
newest version of red cal paper

%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%
\usepackage{subfigure}
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{graphicx}
%\usepackage{algorithm}
%\usepackage[noend]{algpseudocode}
\usepackage{mathrsfs}

@@

%\title[Complex redundant calibration]{Redundant calibration as a complex optimization problem}
\title[Redundant interferometric calibration as a complex optimization problem]{Redundant calibration as a complex optimization problem}
%\author[T.L.~Grobler et al.]{T.L.~Grobler$^{1}$\thanks{E-mail: email@address (AVR)}, G.~Bernardi$^{1,2}$, Z.S.~Ali$^{3}$?, C.L.~Carilli$^{4,5}$?, J.S.~Dillon$^{3}$?, A.~Liu$^{3}$?, \newauthor
%A.R.~Parsons$^{3,6}$? and O.M.~Smirnov$^{1,2}$?\\
%$^{1}$Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
%$^{2}$SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
%$^{3}$Dept. of Astronomy, University of California, Berkeley, CA 94720, USA\\
%$^{4}$National Radio Astronomy Obs., Socorro NM\\
%$^{5}$Cavendish Lab., Cambridge UK\\
%$^{6}$Radio Astronomy Lab., U. California, Berkeley CA 94720, USA}
\author[T.L.~Grobler et al.]{T.L.~Grobler$^{1}$\thanks{E-mail: email@address (AVR)}, G.~Bernardi$^{1,2}$, C.L.~Carilli$^{3,4}$, J.S.~Kenyon$^{1}$, A.R.~Parsons$^{5,6}$? \newauthor and O.M.~Smirnov$^{1,2}$\\
$^{1}$Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa {\bf (GB: add your Stellenbosch affilitation)}\\
$^{2}$SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
$^{3}$National Radio Astronomy Obs., Socorro NM\\
$^{4}$Cavendish Lab., Cambridge UK\\
$^{5}$Dept. of Astronomy, University of California, Berkeley, CA 94720, USA\\
$^{6}$Radio Astronomy Lab., U. California, Berkeley CA 94720, USA}
\begin{document}

@@

\begin{abstract}
Observations of the redshifted 21-cm line from the epoch of reionization have recently motivated the construction of low frequency radio arrays with highly redundant configurations that provide an alternative calibration strategy - ''redundant calibration" - and boosts sensitivity on specific spatial scales. In this paper we formulate calibration of redundant interferometric arrays as a complex optimization problem \citep[i.e.,][]{Smirnov2015} and derive the corresponding calibration equation which we then solve using the Levenber-Marquardt algorithm. This calibration approach is more robust to initial condition than current algorithms and, leveraging on an approximate matrix inversion (``redundant \textsc{StEfCal}"), allows for a very efficient implementation. We investigated the use of the preconditioned conjugate gradient method to solve for the calibration equation without approximating the matrix inversion and found that its computational performance is not competitive with respect to ``redundant \textsc{StEfCal}". The implementation of the new algorithm is made publicly available.
%The construction of new radio interferometers, particularly operating at low frequencies and In this paper we show that instead of splitting our redundant calibration problem into its phase-and-amplitude components (the more traditional approach) we can treat the complex unknown model parameters and their complex conjugates as separate unknowns. We also derive the Levenberg-Maquardt and Gauss-Newton update steps associated with this alternative parameter-and-conjugate framework. We focus more on the Levenberg-Maquardt algorithm in this paper as it exhibits better convergence properties when compared to Gauss-Newton. The Gauss-Newton algorithm is currently the de facto standard that is being used to perform redundant calibration. It is common practice to implement redundant calibration via a hierarchical algorithmic strategy. Using the Levenberg-Maquardt algorithm, therefore, reduces the number of algorithmic layers that are needed to perform redundant calibration. We also compare two previously proposed methods to accelerate redundant calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating direction implicit method outperforms the preconditioned conjugate gradient method. However, the preconditioned conjugate gradient method is more robust as it is more straightforward to adapt to other calibration use cases. We make use of the parameter-and-conjugate framework in this paper, rather than the more traditional amplitude-and-phase approach as it makes it trivial to relate the alternating direction implicit method and the Levenberg-Marquardt algorithm with one another.
\end{abstract}

@@

\section{Introduction}
The quest for the redshifted 21-cm line from the Epoch of Reionization (EoR) is a frontier of modern observational cosmology and has motivated the construction of a series of new interferometric arrays operating at low frequencies over the last decade. EoR measurements are challenging because of the intrinsic faintness of the cosmological signal \citep[see, for instance,][for recent reviews]{Furlanetto2016,McQuinn2016} buried underneath foreground emission which is a few orders of magnitude brighter than the EoR anywhere in the sky \citep[e.g.,][]{Bernardi2009,Bernardi2010,Ghosh2012,Dillon2014,Parsons2014}. 
%and that can be separated by leveraging only on the different spectral properties (citations needed!!!). 
As miscalibrated foreground emission leads to artifacts that can jeopardize the EoR signal \citep[e.g.,][]{grobler2014,barry2016,ewall-wice2016}, an exquisite interferometric calibration is required. Such calibration needs an accurate knowledge of both the sky emission and the instrumental response \citep[e.g.,][]{Smirnov2011c}, unless arrays are deployed in a redundant configuration, i.e. with receiving elements placed on regular grids so that they measure many times the same sky brightness emission. Leveraging on this property, redundancy is a promising path to achieve high accurate interferometric calibration \citep[][]{Noordam1982,Wieringa1992,Pearson1982,Liu2010,Noorishad2012,Marthi2014,Sievers2017}. Moreover, redundant configurations provide a sensitivity boost on a selected sample of spatial scales that can be tuned to be the highest signal--to--noise ratio EoR modes \citep[][]{Parsons2012,Dillon2016}. These reasons motivated the design and deploment of EoR arrays in redundant configurations like the MIT-EoR \citep{Zheng2014}, the Precision Array to Probe the Epoch of Reionization \citep[PAPER,][]{Ali2015} and the Hydrogen Epoch of Reionization Array \citep[HERA,][]{deboer2017}.

@@

In a radio interferometer, the true sky visibilities $y_{pq}$ measured by a baseline formed by antenna $p$ and $q$ are always ``corrupted" by the non-ideal response of the receiver, which is often incorporated into a single, receiver-based complex number $g$. The observed visibility $d_{pq}$ is therefore given by \citep{ME1,ME2,RRIME1}
\begin{equation}
\label{eq:vis_definition}
d_{pq} = g_{p}\conj{g_q} \, y_{pq} + n_{pq},
\end{equation}
where $\conj{(*)}$ indicates conplex conjugation and $n_{pq}$ is the thermal noise component. 
The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a
standard deviation
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}},
\end{equation}
where $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

@@

Radio interferometric calibration aims to determine the best estimate of $\bg = [g_1,g_2,\cdots,g_N]^T$ in order to correct the data and, following equation~\ref{eq:vis_linear_definition}, can be formulated as a non linear least-squares optimization problem:
\begin{equation}
\label{eq:least_squares}
\min_{\bg} \Lambda(\bg) = \min_{\bg} \|\br\|_F^2 = \min_{\bg} \|\bd - \bv(\bg)\|_F^2, 
\end{equation}
where $\Lambda$ is the objective function, $\br$ is the residual vector and $\|\cdot\|$ denotes the Frobenius norm. In standard interferometric calibration, $y_{pq}$ is assumed to be known at some level, for instance through the observation of previously known calibration sources. The knowledge of $y_{pq}$ is often named ``sky model".

@@

Equation~\ref{eq:GN_update_skymodel} or~\ref{eq:LM_update_skymodel} can now be used iteratively to update the parameter vector $\breve{\bg}$:
\begin{equation}
\label{eq:update_skymodel}
\breve{\bg}_{k+1} = \breve{\bg}_{k} + \Delta \breve{\bg}_{k}. 
\end{equation}
until convergence is reached.

@@

Interestingly enough, if $\lambda = 0$ we obtain the odd parameter update step of \textsc{StEfCal}\footnote{$k\in\{0,2,\cdots\}$}, and if $\lambda=1$ (which corresponds
to $\alpha=\frac{1}{2}$) we obtain the even parameter update step of \textsc{StEfCal}\footnote{$k\in\{1,3,\cdots\}$} \citep[\textsc{StEfCal},][]{Mitchell:MWA-cal,Salvini2014}. In the \textsc{StEfCal} algorithm, the measurement equation, equation~\ref{eq:vis_definition}, is linearized by assuming that the gains are known, but that their conjugates are not. Under this assumption, the system of equations become linear and the conjugates of the gains can be obtained straightforwardly. Starting from the latest value of the gain conjugates, an updated estimate of the gains can be obtained and the succession continues until convergence is reached.
Alternating between solving and fixing different sets of parameters (which is exactly what \textsc{StEFCal} does) is in general referred to 
as the alternating direction implicit method (ADI). The \textsc{StEFCal} algorithm reduces the computational complexity of calibration from $\mathbb{O}(N^3)$ to $\mathbb{O}(N^2)$.

\section{Redundant Wirtinger Calibration}
\label{sec:red_wirtinger}
Interferometric baselines are named redundant when they sample the exact same visibilities in the $uv$-plane, i.e. if baseline $pq$ and $rs$ are redundant then $y_{pq} = y_{rs}$. It is convenient to group redundant visibilities together and label each group using a single index rather than using their antenna pairs as in equation~\ref{eq:vis_definition}. We introduce a function $\phi_{pq}$ that maps the antenna pair associated with a specific baseline to the single group index of the redundant baseline group that includes the baseline in question, i.e. if baseline $pq$ and $rs$ are redundant then $\phi_{pq} = \phi_{rs}$. The function $\phi_{pq}$ is not symmetric as $\phi_{pq} = 0$ if $p \geq q$. Equation~\ref{eq:vis_definition} can be re-written for a redundant array as:
\begin{equation}
\label{eq:vis_red}
d_{pq} = g_{p}\conj{g_q}y_{\phi_{pq}} + n_{pq},
\end{equation}
with the same vector form as equation~\ref{eq:vis_red} if
\begin{align}
 \left [ \bd \right]_{\alpha_{pq}} &= d_{pq}, & \left [ \bv \right ]_{\alpha_{pq}} &= v_{pq}=g_p y_{\phi_{pq}} \conj{g_q},\nonumber\\
 \left [ \bn \right ]_{\alpha_{pq}} &= n_{pq}, &  &\label{eq:vec_definitions}
\end{align}
where the vectors in Equation~\ref{eq:vec_definitions} are column vectors of size $B$ (i.e. $p<q$).

@@

We can now formulate redundant calibration as a least-squares problem:
\begin{equation}
\label{eq:least_squares_red}
\min_{\bz} \Lambda(\bz) = \min_{\bz} \|\br\|_F^2 = \min_{\bz} \|\bd - \bv(\bz)\|_F^2, 
\end{equation}
where
\begin{align}
 \bg &=[g_1,\cdots,g_N]^T, & \by &= [y_1,\cdots,y_L]^T,\nonumber\\
 \bz &= [\bg^T,\by^T]^T, &  &\label{eq:parm_definitions}
 \end{align}
where the number of model parameters to be solved for is now $P = 2(N+L)$, since redundant calibration is a complex problem.

In literature, equation~\ref{eq:least_squares_red} is solved by splitting the problem into its real and imaginary parts. The real and imaginary parts of the unknown parameters are then solved separately \citep{Wieringa1992,Liu2010,Zheng2014}. 
Currently, the above is achieved by using the real-valued GN algorithm \citep{Kurien2016}. 
We, instead, intend to formulate the redundant calibration problem using Wirtinger calculus and recast equation~\ref{eq:least_squares_red} as
\begin{equation}
\label{eq:least_squares_complex}
\min_{\breve{\bz}} \|\breve{\br}\| = \min_{\breve{\bz}} \|\breve{\bd} - \breve{\bv}(\breve{\bz})\|, 
\end{equation}
where $\breve{\bz} = [\bz^T,\conj{\bz}^T]^T$. 
We derive the complex Jacobian associated with equation~\ref{eq:least_squares_complex} to be:
\begin{equation}
\label{eq:Jacobian}
\bJ = \begin{bmatrix}
       \bmJ & \bmJ^*\\
       \conj{\bmJ}^* & \conj{\bmJ} 
      \end{bmatrix},
\end{equation}
where 
\begin{equation}
[\bmJ]_{\alpha_{pq},i} = \begin{cases} 
     \frac{\partial v_{pq}}{\partial g_i} & \textrm{if}~i\leq N \\
     \frac{\partial v_{pq}}{\partial y_{i-N}} & \textrm{otherwise}  
\end{cases}, %, & \bmJ^* &= \frac{\partial \bv}{\partial \conj{\bz}}. 
\end{equation}
and
\begin{equation}
[\bmJ^*]_{\alpha_{pq},i} = \begin{cases} 
     \frac{\partial v_{pq}}{\partial \conj{g}_i} & \textrm{if}~i\leq N \\
     \frac{\partial v_{pq}}{\partial \conj{y}_{i-N}} & \textrm{otherwise}  
\end{cases}. %, & \bmJ^* &= \frac{\partial \bv}{\partial \conj{\bz}}. 
\end{equation}
We can now calculate the GN and LM updates to be 
\begin{equation}
\label{eq:GN_update}
\Delta \breve{\bz} = (\bJ^H\bJ)^{-1}\bJ^H\breve{\br}
\end{equation}
and 
\begin{equation}
\label{eq:LM_update}
\Delta \breve{\bz} = (\bJ^H\bJ + \lambda\bD)^{-1}\bJ^H\breve{\br},
\end{equation}
respectively. As in Section~\ref{sec:sky_wirtinger}, equation~\ref{eq:GN_update} can be used to iteratively update our parameter vector:
\begin{equation}
\label{eq:update}
\breve{\bz}_{k+1} = \breve{\bz}_{k} + \Delta \breve{\bz}_{k}. 
\end{equation}
The analytic expressions for $\bJ$, $\bH$ and $\bJ^H\breve{\br}$ can be found in Appendix~\ref{sec:analytic}.
Appendix~\ref{sec:analytic} also contain two useful identities, involving $\bJ$ and $\bH$.

@@

The simulated visibilities included a noise term. The purpose of which is to produce a certain signal--to-noise ratio (SNR) in the simulated
visibilities. We made use of the following definition of \citep[SNR, ][]{Liu2010,Marthi2014}:  
\begin{equation}
\label{}
\textrm{SNR} = \frac{<\bv\odot\conj{\bv}>_{\nu,t,pq}}{<\bn\odot\conj{\bn}>_{\nu,t,pq}}, 
\end{equation}
where $<*>_{\nu,t,pq}$ denotes averaging over frequency, time and baseline.
{\bf (GB; if $n$ is the standard noise term, is the denominator just the noise variance? We need to specify a little more here...)}.
{\bf (TLG; Yes it is the power in the noise. The average power in signal divided by the average noise power. Recall, I confirmed this 
expression with Adrian Liu. I do not know what you wish to add here. I am just trying to say that my simulations are proper, I do not use a relative quantity to 
conduct them. This is the same SNR definition as was used in previous redundant cal papers.)}.

Figure~\ref{fig:prec_error} show the simulation results as a function of SNR and number of antennas. The accuracy of our solutions is quantified through the percentage error
\begin{equation}
\label{eq:beta}
\beta = \frac{\|\bv - \widehat{\bv}\|_F^2}{\|\bv\|_F^2},
\end{equation}
where $\widehat{\bv}$ is the redundant \textsc{StEFCal} parameter estimate.

@@

\citet{Marthi2014} conducted an even more in depth analysis than the one presented here. They found that the total computational cost of redundant StEfCal
is $\mathbb{O}(P^2)$ and that it approaches the Cram{\'e}r-Rao Bound (CRB) at moderate SNRs. 
{\bf (GB: is there a similar analysis in \cite{Liu2010} a/o \cite{Marthi2014}? If so, what do they find?)}.
{\bf (TLG: added.)}

\begin{figure}
\includegraphics[width=0.47\textwidth]{./prec_error.pdf} 
\caption{We plot the percentage error $\beta$ between the simulated visibilitied and the visibilities solved for by \textsc{StEFCal} for different SNR values as a function of the number of antennas in the array $N$. {\bf (GB: I don't undesrtand the curve with SNR = -1... The yaxis should read $\beta$).}}
\label{fig:prec_error}
\end{figure}

\section{Preconditioned Conjugate Gradient Method}
\label{sec:pcg}
\citet{Liu2010} suggested that the speed of convergence of redundant calibration could be optimized using the 
conjugate gradient method \citep{Hestenes1952}, which would be computationally advantageous, since the Hessian matrix 
is sparse \citep{Reid1971}. In this section we study the computational complexity
of the conjugate gradient method when used to invert the Hessian matrix in equation~\ref{eq:LM_update}, 
in particular through the use of preconditioning (i.e. the preconditioned conjugate gradient (PCG) method). 
%(see also Appendix~\ref{sec:conj_grad}).

>>>>

%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%
\usepackage{subfigure}
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{graphicx}
%\usepackage{algorithm}
%\usepackage[noend]{algpseudocode}
\usepackage{mathrsfs}
\usepackage{url}

@@

%\title[Complex redundant calibration]{Redundant calibration as a complex optimization problem}
\title[Redundant interferometric calibration as a complex optimization problem]{Redundant calibration as a complex optimization problem}
%\author[T.L.~Grobler et al.]{T.L.~Grobler$^{1}$\thanks{E-mail: email@address (AVR)}, G.~Bernardi$^{1,2}$, Z.S.~Ali$^{3}$?, C.L.~Carilli$^{4,5}$?, J.S.~Dillon$^{3}$?, A.~Liu$^{3}$?, \newauthor
%A.R.~Parsons$^{3,6}$? and O.M.~Smirnov$^{1,2}$?\\
%$^{1}$Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
%$^{2}$SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
%$^{3}$Dept. of Astronomy, University of California, Berkeley, CA 94720, USA\\
%$^{4}$National Radio Astronomy Obs., Socorro NM\\
%$^{5}$Cavendish Lab., Cambridge UK\\
%$^{6}$Radio Astronomy Lab., U. California, Berkeley CA 94720, USA}
\author[T.L.~Grobler et al.]{T.L.~Grobler$^{1}$\thanks{E-mail: email@address (AVR)}, G.~Bernardi$^{2,3}$, C.L.~Carilli$^{4,5}$, J.S.~Kenyon$^{2}$, A.R.~Parsons$^{6,7}$? \newauthor and O.M.~Smirnov$^{2,3}$\\
$^{1}$Dept of Mathematical Sciences, Computer Science Division, Stellenbosch University, Private Bag X1, 7602 Matieland, South Africa\\
$^{2}$Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
$^{3}$SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
$^{4}$National Radio Astronomy Obs., Socorro NM\\
$^{5}$Cavendish Lab., Cambridge UK\\
$^{6}$Dept. of Astronomy, University of California, Berkeley, CA 94720, USA\\
$^{7}$Radio Astronomy Lab., U. California, Berkeley CA 94720, USA}
\begin{document}

@@

\begin{abstract}
Observations of the redshifted 21-cm line from the epoch of reionization have recently motivated the construction of low 
frequency radio arrays with highly redundant configurations that provide an alternative calibration strategy - 
''redundant calibration" - and boosts sensitivity on specific spatial scales. In this paper we formulate 
calibration of redundant interferometric arrays as a complex optimization problem \citep[i.e.,][]{Smirnov2015}. 
We then solve this optimization problem via the Levenberg-Marquardt algorithm. This calibration approach is more 
robust to initial conditions than current algorithms and, leveraging on an approximate matrix inversion, allows for further optimization and an efficient implementation (``redundant \textsc{StEfCal}"). We also investigated the use of the preconditioned conjugate gradient method as an alternative optimization strategy, instead of 
an approximate matrix inverse, but found that its computational performance is not competitive with respect to ``redundant \textsc{StEfCal}". The efficient implementation of this new algorithm is made publicly available.
%The construction of new radio interferometers, particularly operating at low frequencies and In this paper we show that instead of splitting our redundant calibration problem into its phase-and-amplitude components (the more traditional approach) we can treat the complex unknown model parameters and their complex conjugates as separate unknowns. We also derive the Levenberg-Maquardt and Gauss-Newton update steps associated with this alternative parameter-and-conjugate framework. We focus more on the Levenberg-Maquardt algorithm in this paper as it exhibits better convergence properties when compared to Gauss-Newton. The Gauss-Newton algorithm is currently the de facto standard that is being used to perform redundant calibration. It is common practice to implement redundant calibration via a hierarchical algorithmic strategy. Using the Levenberg-Maquardt algorithm, therefore, reduces the number of algorithmic layers that are needed to perform redundant calibration. We also compare two previously proposed methods to accelerate redundant calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating direction implicit method outperforms the preconditioned conjugate gradient method. However, the preconditioned conjugate gradient method is more robust as it is more straightforward to adapt to other calibration use cases. We make use of the parameter-and-conjugate framework in this paper, rather than the more traditional amplitude-and-phase approach as it makes it trivial to relate the alternating direction implicit method and the Levenberg-Marquardt algorithm with one another.
\end{abstract}

@@

\section{Introduction}
The quest for the redshifted 21-cm line from the Epoch of Reionization (EoR) is a frontier of modern observational cosmology and has motivated the construction of a series of new interferometric arrays operating at low frequencies over the last decade. EoR measurements are challenging because of the intrinsic faintness of the cosmological signal \citep[see, for instance,][for recent reviews]{Furlanetto2016,McQuinn2016} buried underneath foreground emission which is a few orders of magnitude brighter than the EoR anywhere in the sky \citep[e.g.,][]{Bernardi2009,Bernardi2010,Ghosh2012,Dillon2014,Parsons2014}. 
%and that can be separated by leveraging only on the different spectral properties (citations needed!!!). 
As miscalibrated foreground emission leads to artifacts that can jeopardize the EoR signal \citep[e.g.,][]{grobler2014,barry2016,ewall-wice2016}, an exquisite interferometric calibration is required. Such calibration needs an accurate knowledge of both the sky emission and the instrumental response \citep[e.g.,][]{Smirnov2011c}, unless arrays are deployed in a redundant configuration, i.e. with receiving elements placed on regular grids so that they measure many times the same sky brightness emission. Leveraging on this property, redundancy is a promising path to achieve high accurate interferometric calibration \citep[][]{Noordam1982,Wieringa1992,Pearson1982,Liu2010,Noorishad2012,Marthi2014,Sievers2017}. Moreover, redundant configurations provide a sensitivity boost on a selected sample of spatial scales that can be tuned to be the highest signal--to--noise ratio (SNR) EoR modes \citep[][]{Parsons2012,Dillon2016}. These reasons motivated the design and deploment of EoR arrays in redundant configurations like the MIT-EoR \citep{Zheng2014}, the Precision Array to Probe the Epoch of Reionization \citep[PAPER,][]{Ali2015} and the Hydrogen Epoch of Reionization Array \citep[HERA,][]{deboer2017}.

@@

In a radio interferometer, the true sky visibilities $y_{pq}$ measured by a baseline formed by antenna $p$ and $q$ are always ``corrupted" by the non-ideal response of the receiver, which is often incorporated into a single, receiver-based complex number $g$. The observed visibility $d_{pq}$ is therefore given by \citep{ME1,ME2,RRIME1}
\begin{equation}
\label{eq:vis_definition}
d_{pq} = g_{p}\conj{g_q} \, y_{pq} + n_{pq},
\end{equation}
where $\conj{(*)}$ indicates conplex conjugation and $n_{pq}$ is the thermal noise component. 
The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a
standard deviation $\sigma$:
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}},
\end{equation}
where $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

@@

Radio interferometric calibration aims to determine the best estimate of $\bg = [g_1,g_2,\cdots,g_N]^T$ in order to correct the data and, following equation~\ref{eq:vis_linear_definition}, can be formulated as a non linear least-squares optimization problem:
\begin{equation}
\label{eq:least_squares}
\min_{\bg} \Lambda(\bg) = \min_{\bg} \|\br\|_F^2 = \min_{\bg} \|\bd - \bv(\bg)\|_F^2, 
\end{equation}
where $\Lambda$ is the objective function, $\br$ is the residual vector and $\|\cdot\|_F$ denotes the Frobenius norm. In standard interferometric calibration, $y_{pq}$ is assumed to be known at some level, for instance through the observation of previously known calibration sources. The knowledge of $y_{pq}$ is often named ``sky model".

@@

Equation~\ref{eq:GN_update_skymodel} or~\ref{eq:LM_update_skymodel} can now be used iteratively to update the parameter vector $\breve{\bg}$:
\begin{equation}
\label{eq:update_skymodel}
\breve{\bg}_{k+1} = \breve{\bg}_{k} + \Delta \breve{\bg}_{k},
\end{equation}
until convergence is reached.

@@

Interestingly enough, if $\lambda = 0$ we obtain the odd parameter update step of \textsc{StEfCal}\footnote{$k\in\{0,2,\cdots\}$}, and if $\lambda=1$ (which corresponds
to $\alpha=\frac{1}{2}$) we obtain the even parameter update step of \textsc{StEfCal}\footnote{$k\in\{1,3,\cdots\}$} \citep[\textsc{StEfCal},][]{Mitchell:MWA-cal,Salvini2014}. In the \textsc{StEfCal} algorithm, the measurement equation, equation~\ref{eq:vis_definition}, is linearized by assuming that the gains are known, but that their conjugates are not. Under this assumption, the system of equations become linear and the conjugates of the gains can be obtained straightforwardly. Starting from the latest value of the gain conjugates, an updated estimate of the gains can be obtained and the succession continues until convergence is reached.
Alternating between solving and fixing different sets of parameters (which is exactly what \textsc{StEfCal} does) is in general referred to 
as the alternating direction implicit method (ADI). The \textsc{StEfCal} algorithm reduces the computational complexity of calibration from $O(N^3)$ to $O(N^2)$.

\section{Redundant Wirtinger Calibration}
\label{sec:red_wirtinger}
Interferometric baselines are named redundant when they sample the exact same visibilities in the $uv$-plane, i.e. if baseline $pq$ and $rs$ are redundant then $y_{pq} = y_{rs}$. It is convenient to group redundant visibilities together and label each group using a single index rather than using their antenna pairs as in equation~\ref{eq:vis_definition}. We introduce a function $\phi_{pq}$ that maps the antenna pair associated with a specific baseline to the single group index of the redundant baseline group that includes the baseline in question, i.e. if baseline $pq$ and $rs$ are redundant then $\phi_{pq} = \phi_{rs}$. The function $\phi_{pq}$ is not symmetric as $\phi_{pq} = 0$ if $p \geq q$. Equation~\ref{eq:vis_definition} can be re-written for a redundant array as:
\begin{equation}
\label{eq:vis_red}
d_{pq} = g_{p}\conj{g_q}y_{\phi_{pq}} + n_{pq},
\end{equation}
with the same vector form as equation~\ref{eq:vis_linear_definition} if
\begin{align}
 \left [ \bd \right]_{\alpha_{pq}} &= d_{pq}, & \left [ \bv \right ]_{\alpha_{pq}} &= v_{pq}=g_p y_{\phi_{pq}} \conj{g_q},\nonumber\\
 \left [ \bn \right ]_{\alpha_{pq}} &= n_{pq}, &  &\label{eq:vec_definitions}
\end{align}
where the vectors in Equation~\ref{eq:vec_definitions} are column vectors of size $B$ (i.e. $p<q$).

@@

We can now formulate redundant calibration as a least-squares problem:
\begin{equation}
\label{eq:least_squares_red}
\min_{\bz} \Lambda(\bz) = \min_{\bz} \|\br\|_F^2 = \min_{\bz} \|\bd - \bv(\bz)\|_F^2, 
\end{equation}
where
\begin{align}
 \bg &=[g_1,\cdots,g_N]^T, & \by &= [y_1,\cdots,y_L]^T,\nonumber\\
 \bz &= [\bg^T,\by^T]^T. &  &\label{eq:parm_definitions}
 \end{align}
The number of model parameters to be solved for is now $P = 2(N+L)$, since redundant calibration is a complex problem.

In literature, equation~\ref{eq:least_squares_red} is solved by splitting the problem into its real and imaginary parts. The real and imaginary parts of the unknown parameters are then solved for separately \citep{Wieringa1992,Liu2010,Zheng2014}. 
Currently, the above is achieved by using the real-valued GN algorithm \citep{Kurien2016}. 
We, instead, intend to formulate the redundant calibration problem using Wirtinger calculus and recast equation~\ref{eq:least_squares_red} as
\begin{equation}
\label{eq:least_squares_complex}
\min_{\breve{\bz}} \|\breve{\br}\| = \min_{\breve{\bz}} \|\breve{\bd} - \breve{\bv}(\breve{\bz})\|, 
\end{equation}
where $\breve{\bz} = [\bz^T,\conj{\bz}^T]^T$. 
We derive the complex Jacobian associated with equation~\ref{eq:least_squares_complex} to be:
\begin{equation}
\label{eq:Jacobian}
\bJ = \begin{bmatrix}
       \bmJ & \bmJ^*\\
       \conj{\bmJ}^* & \conj{\bmJ} 
      \end{bmatrix},
\end{equation}
where 
\begin{equation}
[\bmJ]_{\alpha_{pq},i} = \begin{cases} 
     \frac{\partial v_{pq}}{\partial g_i} & \textrm{if}~i\leq N \\
     \frac{\partial v_{pq}}{\partial y_{i-N}} & \textrm{otherwise}  
\end{cases}, %, & \bmJ^* &= \frac{\partial \bv}{\partial \conj{\bz}}. 
\end{equation}
and
\begin{equation}
[\bmJ^*]_{\alpha_{pq},i} = \begin{cases} 
     \frac{\partial v_{pq}}{\partial \conj{g}_i} & \textrm{if}~i\leq N \\
     \frac{\partial v_{pq}}{\partial \conj{y}_{i-N}} & \textrm{otherwise}  
\end{cases}. %, & \bmJ^* &= \frac{\partial \bv}{\partial \conj{\bz}}. 
\end{equation}
We can now calculate the GN and LM updates to be 
\begin{equation}
\label{eq:GN_update}
\Delta \breve{\bz} = (\bJ^H\bJ)^{-1}\bJ^H\breve{\br}
\end{equation}
and 
\begin{equation}
\label{eq:LM_update}
\Delta \breve{\bz} = (\bJ^H\bJ + \lambda\bD)^{-1}\bJ^H\breve{\br},
\end{equation}
respectively. As in Section~\ref{sec:sky_wirtinger}, equation~\ref{eq:GN_update} can be used to iteratively update our parameter vector:
\begin{equation}
\label{eq:update}
\breve{\bz}_{k+1} = \breve{\bz}_{k} + \Delta \breve{\bz}_{k}. 
\end{equation}
The analytic expressions for $\bJ$, $\bH$ and $\bJ^H\breve{\br}$ can be found in Appendix~\ref{sec:analytic}.
Appendix~\ref{sec:analytic} also contain two useful identities, involving $\bJ$ and $\bH$.

@@

The simulated visibilities included a noise term. The purpose of which is to produce a certain SNR in the simulated
visibilities. We made use of the following definition of \citep[SNR, ][]{Liu2010,Marthi2014}:  
\begin{equation}
\label{}
\textrm{SNR} = 10\log \left (\frac{<\bv\odot\conj{\bv}>_{\nu,t,pq}}{<\bn\odot\conj{\bn}>_{\nu,t,pq}} \right ), 
\end{equation}
where $<*>_{\nu,t,pq}$ denotes averaging over frequency, time and baseline.
{\bf (GB; if $n$ is the standard noise term, is the denominator just the noise variance? We need to specify a little more here...)}.
{\bf (TLG; Yes it is the power in the noise. The average power in signal divided by the average noise power. Recall, I confirmed this 
expression with Adrian Liu. I do not know what you wish to add here. I am just trying to say that my simulations are proper, I do not use a relative quantity to 
conduct them. This is the same SNR definition as was used in previous redundant cal papers)}.

Figure~\ref{fig:prec_error} show the simulation results as a function of SNR and number of antennas. The accuracy of our solutions is quantified through the percentage error
\begin{equation}
\label{eq:beta}
\beta = \frac{\|\bv - \widehat{\bv}\|_F^2}{\|\bv\|_F^2},
\end{equation}
where $\widehat{\bv}$ is the redundant \textsc{StEfCal} parameter estimate.

@@

\citet{Marthi2014} conducted an even more in depth analysis than the one presented here. They found that the total computational cost of the algorithm associated with redundant StEfCal
is $O(P^2)$ and that it approaches the Cram{\'e}r-Rao Bound (CRB) at moderate SNRs. 
{\bf (GB: is there a similar analysis in \cite{Liu2010} a/o \cite{Marthi2014}? If so, what do they find?)}.
{\bf (TLG: Added.)}

\begin{figure}
\includegraphics[width=0.47\textwidth]{./prec_error.pdf} 
\caption{We plot the percentage error $\beta$ between the simulated visibilitied and the visibilities solved for by redundant \textsc{StEfCal} for different SNR values as a function of the number of antennas in the array $N$. {\bf (GB: I don't undesrtand the curve with SNR = -1... The yaxis should read $\beta$).} {\bf (TLG: Added $\beta$, SNR is expressed in dB. Made this more explicit).}}
\label{fig:prec_error}
\end{figure}

\section{Preconditioned Conjugate Gradient Method}
\label{sec:pcg}
\citet{Liu2010} suggested that the speed of convergence of redundant calibration could be optimized using the 
conjugate gradient method \citep{Hestenes1952}, which would be computationally advantageous, since the Hessian matrix associated 
with redundant calibration (see Figure~\ref{fig:hessian}) is sparse \citep{Reid1971}. In this section we study the computational complexity
of the conjugate gradient (CG) method when it is used to invert the modified Hessian matrix (see equation~\ref{eq:LM_update}), 
in particular when preconditioning is used (i.e. the PCG method). Interestingly, emperical tests suggests that the unmodified Hessian itself is singular. It is therefore important to mention here, that the CG method 
can pseudo invert the unmodified Hessian as well, i.e. the CG method can be directly applied to equation~\ref{eq:GN_update}, since the Hessian is a positive semi-definite Hermition matrix and the vector $\bJ^H\breve{\br}$ is an element of its column range \citep{Lu2015}.

%(see also Appendix~\ref{sec:conj_grad}).
